The Success of College Basketball Teams can be predicted using several advanced team statistics

### 2022-04-23 Paper-On-NCAAB-Stats/outputs/paper/Paper-On-NCAAB-Stats.Rmd
prelim results section

All the variables to the right of SRS in Table \@ref(tab:dataextract) are the team statistics that can potentially be explanatory variable in the model. The team statistics included in the model will only be the most important ones that clearly play a role in making a team better or more successful. There were additional statistics in the original dataset, but they were removed as they did not reasonably seem like they had an impact on how well a team did, so they were removed. The remaining statistics in order are as follows; pace, offensive rating, 3-point attempt rate, true shooting percentage, total rebound percentage, assist percentage, steal percentage, block percentage, turnover percentage, and offensive rebound percentage. Once only the main and important statistics from this list are chosen they will be defined and looked at further, but for now we will focus on analyzing our response variable and its relationship with these team statistics.

@@

So, none of these 6 variables will be considered for the model, and only offensive rating, true shooting percentage, total rebound percentage, and turnover percentage will potentially be used to predict simple rating system, our response variable, and help us determine what leads to team success.

@@

ncaab_stats %>% 
  select(srs, o_rtg, ts_percent, trb_percent, tov_percent) %>% 
  datasummary_skim(histogram=FALSE, output = "kableExtra", 
                   title="Summary statistics of important variables") %>% 
  kable_styling(latex_options= "hold_position")
```
Table \@ref(tab:summarytab) shows several summary statistics of each of the statistics that will be considered for the model, including the response. Unique shows the number of different unique values for each team statistic. Missing shows the percent of values missing in the data for each variable, and each one is 0 since all NA values were removed in the initial cleaning process of the dataset. The noteworthy statistics included in the table are mean, standard deviation, mininum, median, and maximum. We once again see how simple rating system has a mean and median quite close together and is centered very close to one, showing how the distribution is only slightly skewed and is close to being a normal distribution. With a standard deviation of just over 10, there is a decent amount of spread in the distribution but nothing too substantial. The larger max and min may cause some impliations in the future but only minimally.

All the potential explanatory team statistics seem to have little to no skew at all as the mean and median of each variable is exactly the same, as they are 0.5 for true shooting percentage, 50.3 for total rebound percentage and 16.2 for turnover percentage, except for offensive rating whose mean and median only differ by 0.1 (102.9 and 103). Each variable also seems to have fairly reasonably low standard deviations, meaning no variable has a spread too high that will affect our model. Based on the maximum's and minimum's, there does not seem to be any crazy high or low values that are serious outliers and would skew our results too drastically either, so overall the data looks quite good.

@@

```{r dags, echo=FALSE, fig.cap="A DAG showing Offensive Rating as a mediator}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

@@

$$\hat{y_i} = \hat{\beta_{0}} + \hat{\beta_{1}}x_1 + \hat{\beta_{2}}x_2 + \hat{\beta_{3}}x_3 + {\epsilon} $$
where:
- $\hat{y_i}$ represents the predicted value of SRS given specific values of the predictor variables
- $\hat{\beta_{0}}$ is the estimated intercept coefficient
- $x_1$ represents the predictor variable of True Shooting Percentage
- $\hat{\beta_{1}}$ is the estimated slope coefficient for True Shooting Percentage
- $x_2$ represents the predictor variable of Total Rebound Percentage
- $\hat{\beta_{2}}$ is the estimated slope coefficient for Total Rebound Percentage
- $x_3$ represents the predictor variable of Turnover Percentage
- $\hat{\beta_{3}}$ is the estimated slope coefficient for Turnover Percentage
- and $\epsilon$ is an error term

@@

```{r}
fitted_and_resids <- 
  augment(mod_ts_trb_tov,
          data = ncaab_stats_train)

@@

```{r assumplots, fig.height=8, fig.cap="Residuals and Fitted Value plots for linear regression model predicting Simple Rating System using a training dataset on college basketball statistics}
ggarrange(assump_plot1,assump_plot5,assump_plot2,assump_plot3, assump_plot4, ncol=2, nrow=3)
```

>>>>

All the variables to the right of SRS in Table \@ref(tab:dataextract) are the numeric team statistics that can potentially be explanatory variable in the model. The team statistics included in the model will only be the most important ones that clearly play a role in making a team better or more successful. There were additional statistics in the original dataset, but they were removed as they did not reasonably seem like they had an impact on how well a team did, so they were removed. The remaining statistics in order are as follows; pace, offensive rating, 3-point attempt rate, true shooting percentage, total rebound percentage, assist percentage, steal percentage, block percentage, turnover percentage, and offensive rebound percentage. Once only the main and important statistics from this list are chosen they will be defined and looked at further, but for now we will focus on analyzing our response variable and its relationship with these team statistics.

@@

So, none of these 6 variables will be considered for the model, and only offensive rating, true shooting percentage, total rebound percentage, and turnover percentage will potentially be used to predict simple rating system, our response variable, and help us determine what leads to team success. Each of these variables are once again numeric, with offensiove rating being a rate per 100 possessions, and the remaining three other variables all being in the form of percentages,

@@

ncaab_stats %>% 
  select(srs, o_rtg, ts_percent, trb_percent, tov_percent) %>% 
  datasummary_skim(histogram=FALSE, output = "kableExtra", 
                   title="Summary statistics of important variables") %>% 
  kable_styling(latex_options= "hold_position")
```
Table \@ref(tab:summarytab) shows several summary statistics of each of these statistics that will be considered for the model, including the response. Unique shows the number of different unique values for each team statistic. Missing shows the percent of values missing in the data for each variable, and each one is 0 since all NA values were removed in the initial cleaning process of the dataset. The noteworthy statistics included in the table are mean, standard deviation, minimum, median, and maximum. We once again see how simple rating system has a mean and median quite close together and is centered very close to one, showing how the distribution is only slightly skewed and is close to being a normal distribution. With a standard deviation of just over 10, there is a decent amount of spread in the distribution but nothing too substantial. The larger max and min may cause some implications in the future but only minimally.

All the potential explanatory team statistics seem to have little to no skew at all as the mean and median of each variable are extremely similar. They are actually the exact same for total rebound percentage, with a mean and median of 50.3, and for turnover percentage, with a mean and median of 16.2. Although they are not identical, the mean and medians for offensive rating and true shooting percentage are nearly the same as well, as they only differ by a tenth for both, as they are 102.9 and 103 respectively for offensive rating, and 53.8 and 53.7 for true shooting percentage. Each variable also seems to have fairly reasonably low standard deviations, meaning no variable has a spread too high that will affect our model. Based on the maximum's and minimum's, there does not seem to be any crazy high or low values that are serious outliers and would skew our results too drastically either, so overall the data looks quite good.

@@

```{r dags, echo=FALSE, fig.cap="A DAG showing Offensive Rating as a mediator"}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

@@

$$\hat{y_i} = \hat{\beta_{0}} + \hat{\beta_{1}}x_1 + \hat{\beta_{2}}x_2 + \hat{\beta_{3}}x_3 + {\epsilon} $$
where:

* _$\hat{y_i}$ represents the predicted value of SRS given specific values of the predictor variables_
* _$\hat{\beta_{0}}$ is the estimated intercept coefficient_
* _$x_1$ represents the predictor variable of True Shooting Percentage_
* _$\hat{\beta_{1}}$ is the estimated slope coefficient for True Shooting Percentage_
* _$x_2$ represents the predictor variable of Total Rebound Percentage_
* _$\hat{\beta_{2}}$ is the estimated slope coefficient for Total Rebound Percentage_
* _$x_3$ represents the predictor variable of Turnover Percentage_
* _$\hat{\beta_{3}}$ is the estimated slope coefficient for Turnover Percentage_
* _and $\epsilon$ is an error term_

@@

```{r coefsum}
final_model <- lm(srs~ts_percent + trb_percent + tov_percent, data=ncaab_stats_train)

coefs <- format(round(final_model$coefficients, 2))
confints_0 <- format(round(confint(final_model),2))
confints <- str_c("(", trimws(confints_0[,1]), ", ", confints_0[,2], ")")
pvals <- format(signif(summary(final_model)$coefficients[,4], 3))

sum_table <- cbind(coefs, confints, pvals)
rownames(sum_table) <- c("Intercept", "True Shooting %", "Total Rebound %", "Turnover %")
colnames(sum_table) <- c("Estimated Coefficient", "95% Confidence Interval", "P-Value")

kable(sum_table,
    caption= "Summary of parameter values and statistics for linear model predicting Simple Rating System using variables True Shooting %, Total Rebound %, and Turnover %",
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c', 'c')
  ) %>% 
kable_styling(latex_options= "hold_position")

Table \@ref(tab:coefsum) displays the results of our models estimated parameter statistics. It displays each of the estimated coefficient values, the 95% confidence interval for each estimated coefficient, and a p-value stating the significance of each variable. The parameters can be interpreted in order for us to better understand the relationship between the three predictor variables used in the model and Simple Rating System, or a college basketballs team success, shown in the model, and this is especially true for the estimated coefficients. The intercept coefficient of -98.62 tells us what a teams SRS will be if all predicted variable values are equal to zero. So, if in a given season a team hits 0% of their shots, grabs 0% of all possible rebounds, and turns the ball over in 0% of their positions, they will have a Simple Rating System value of -98.62. This is a reasonable intercept value as if a team hits no shots they will score no points, and by grabbing no rebounds they will give the other team many chances to score, so the team will lose many games by a large amount, resulting in the predicted very low SRS value. Of course it is near impossible for a team to have a season where they do not make a shot, do not grab a rebound, and never turn the ball over, but this is just a baseline value that helps us understand how the model works.

Next, the slope coefficient for the true shooting percentage variable is 0.92, meaning that when all other predictor variables are held at a fixed value, if true shooting percentage increases by 1% then their simple rating system will increase by 0.92. Total rebound percentage has a slope coefficient even larger at 1.63. This coefficient value of 1.63 can be interpreted as how much a teams simple rating system increases by when their total rebound percentage goes up 1%. Both of these coefficients make sense as simple rating system will logically increase if a team shoots more efficiently and scores more points or if they grab more rebounds and get more attempts to score that their opposition, which was already determined by looking at the relation graphically earlier.

```{r, echo=FALSE}
fitted_and_resids <- 
  augment(final_model,
          data = ncaab_stats_train)

@@

```{r assumplots, fig.height=8, fig.cap="Residuals and Fitted Value plots for linear regression model predicting Simple Rating System using a training dataset on college basketball statistics"}
ggarrange(assump_plot1,assump_plot5,assump_plot2,assump_plot3, assump_plot4, ncol=2, nrow=3)
```
